---
title: "Historical Experimental Data Analysis"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(jsonlite)
library(tidyverse)
library(ggplot2)
```

## Load and Preprocess Data

```{r load-data}
raw_df <- read_csv("../data/experimental_data.csv")

paired_df <- raw_df %>%
  mutate(
    decision_tuple = map2(c, opp_c, ~ {
      p1 <- fromJSON(.x)
      p2 <- fromJSON(.y)
      map2(p1, p2, ~ c(as.integer(.x == 1), as.integer(.y == 1)))
    }),
    payoff_tuple = map2(payoff, opp_c, ~ {
      p1 <- fromJSON(.x)
      p2_decisions <- fromJSON(.y)
      map2(p1, p2_decisions, ~ c(.x, NA))
    })
  )

if (!dir.exists("../figures")) dir.create("../figures")
```

## Data Overview

```{r data-overview}
n_sequences <- nrow(paired_df)
n_participants <- n_distinct(paired_df$id)
n_sessions <- n_distinct(paired_df$session)
n_papers <- n_distinct(paired_df$paper)
n_game_configs <- paired_df %>%
  distinct(g, l, delta) %>%
  nrow()
avg_seq_length <- mean(map_int(paired_df$decision_tuple, length))

tibble(
  Statistic = c(
    "Total sequences",
    "Unique participants",
    "Unique sessions",
    "Source papers",
    "Unique (g, l, δ) combinations",
    "Average sequence length"
  ),
  Value = c(
    scales::comma(n_sequences),
    scales::comma(n_participants),
    scales::comma(n_sessions),
    scales::comma(n_papers),
    scales::comma(n_game_configs),
    sprintf("%.1f rounds", avg_seq_length)
  )
) %>%
  knitr::kable()
```

### Source Papers

```{r papers}
paired_df %>%
  count(paper, name = "N") %>%
  arrange(desc(N)) %>%
  mutate(N = scales::comma(N)) %>%
  knitr::kable()
```

## Compute Delta Metrics

```{r delta-metrics}
paired_df <- paired_df %>%
  mutate(
    delta_spe = g / (1 + g),
    delta_rd  = (g + l) / (1 + g + l),
    Delta_rd  = delta - delta_rd
  )
```

## Distribution of Deltas

```{r histogram-deltas, fig.width=10, fig.height=6}
parsed_labels <- c(
  delta      = "delta",
  delta_spe  = "delta[\"spe\"]",
  delta_rd   = "delta[\"rd\"]",
  Delta_rd   = "Delta[\"rd\"]"
)
fill_colors <- c(
  delta      = "red",
  delta_spe  = "blue",
  delta_rd   = "green",
  Delta_rd   = "purple"
)

paired_df %>%
  select(delta_spe, delta_rd, delta, Delta_rd) %>%
  pivot_longer(everything(), names_to = "measure", values_to = "value") %>%
  mutate(measure = factor(measure, levels = names(parsed_labels))) %>%
  ggplot(aes(x = value, fill = measure)) +
  geom_histogram(bins = 20, color = "black", alpha = 0.5) +
  facet_wrap(~ measure, scales = "free", labeller = as_labeller(parsed_labels, label_parsed)) +
  scale_fill_manual(values = fill_colors) +
  labs(x = NULL, y = "Count", title = "Distribution of δ (and Related Game Metadata)") +
  theme_bw() +
  theme(legend.position = "none")

ggsave("../figures/3b_histograms_deltas.png", width = 10, height = 6)
```

## Distribution of g and l

```{r histogram-gl, fig.width=10, fig.height=4}
gl_hist_df <- paired_df %>%
  select(g, l) %>%
  pivot_longer(cols = everything(), names_to = "component", values_to = "value") %>%
  mutate(component = factor(component, levels = c("g", "l")))

bin_width <- 0.5
x_min <- floor(min(gl_hist_df$value, na.rm = TRUE) / bin_width) * bin_width
x_max <- ceiling(max(gl_hist_df$value, na.rm = TRUE) / bin_width) * bin_width
breaks_seq <- seq(x_min, x_max, by = bin_width)

ggplot(gl_hist_df, aes(x = value, fill = component)) +
  geom_histogram(
    breaks = breaks_seq,
    color = "black",
    alpha = 0.5,
    boundary = 0
  ) +
  facet_wrap(~ component, labeller = label_parsed) +
  labs(
    x = NULL,
    y = "Count",
    title = "Distribution of g and l"
  ) +
  theme_bw() +
  theme(legend.position = "none")

ggsave("../figures/3b_histogram_gl.png", width = 10, height = 4)
```

## Player 1 Cooperation Metrics

```{r p1-coop-metrics}
paired_df <- paired_df %>%
  mutate(
    p1_coop_rate = map_dbl(decision_tuple, ~ mean(map_dbl(.x, 1))),
    p1_r1_coop   = map_int(decision_tuple, ~ .x[[1]][1])
  )
```

## Binned Cooperation Rates by Delta

```{r binned-coop}
delta_vals <- paired_df %>%
  filter(!is.na(delta)) %>%
  group_by(delta) %>%
  summarise(
    bin_center = mean(delta),
    coop_rate  = mean(p1_coop_rate),
    r1_coop    = mean(p1_r1_coop),
    measure = "delta",
    .groups = "drop"
  )

other_binned <- paired_df %>%
  select(p1_coop_rate, p1_r1_coop, delta_spe, delta_rd, Delta_rd) %>%
  pivot_longer(c(delta_spe, delta_rd, Delta_rd), names_to = "measure", values_to = "x") %>%
  group_by(measure) %>%
  mutate(bin = ntile(x, 10)) %>%
  group_by(measure, bin) %>%
  summarise(
    bin_center = mean(x),
    coop_rate  = mean(p1_coop_rate),
    r1_coop    = mean(p1_r1_coop),
    .groups = "drop"
  )

binned_df <- bind_rows(delta_vals, other_binned) %>%
  mutate(measure = factor(measure, levels = names(parsed_labels)))
```

```{r plot-coop-rate, fig.width=10, fig.height=6}
ggplot(binned_df, aes(x = bin_center, y = coop_rate, color = measure)) +
  geom_point(size = 2) +
  facet_wrap(~ measure, scales = "free_x", labeller = as_labeller(parsed_labels, label_parsed)) +
  scale_color_manual(values = fill_colors) +
  labs(x = NULL, y = "P1 Cooperation Rate", title = "P1 Cooperation Rate by δ (and Related Game Metadata)") +
  theme_bw() +
  theme(legend.position = "none")

ggsave("../figures/3b_p1_coop_rate_binned.png", width = 10, height = 6)
```

```{r plot-r1-coop, fig.width=10, fig.height=6}
ggplot(binned_df, aes(x = bin_center, y = r1_coop, color = measure)) +
  geom_point(size = 2) +
  facet_wrap(~ measure, scales = "free_x", labeller = as_labeller(parsed_labels, label_parsed)) +
  scale_color_manual(values = fill_colors) +
  labs(x = NULL, y = "P1 Round 1 Cooperation", title = "P1 First-Round Cooperation by δ (and Related Game Metadata)") +
  theme_bw() +
  theme(legend.position = "none")

ggsave("../figures/3b_p1_r1_coop_binned.png", width = 10, height = 6)
```

## Memory-1 Histories

```{r memory-1}
extract_history <- function(seq) {
  if (length(seq) < 1) {
    return(tibble(previous_pair = character(), next_pair = character(), next_p1 = integer(), round_next = integer()))
  }

  history <- tibble(
    previous_pair = "initial",
    next_pair = paste0(seq[[1]], collapse = ""),
    next_p1 = seq[[1]][1],
    round_next = 1L
  )

  if (length(seq) >= 2) {
    after <- tibble(
      previous_pair = map_chr(seq[-length(seq)], ~ paste0(.x, collapse = "")),
      next_pair     = map_chr(seq[-1], ~ paste0(.x, collapse = "")),
      next_p1       = map_int(seq[-1], ~ .x[1]),
      round_next    = 2L:length(seq)
    )
    history <- bind_rows(history, after)
  }
  return(history)
}

history_df <- paired_df %>%
  mutate(transitions = map(decision_tuple, extract_history)) %>%
  unnest(transitions)

label_binary_pair <- function(pair) {
  recode(pair,
         "11" = "CC", "10" = "CD", "01" = "DC", "00" = "DD",
         "initial" = "initial", .default = pair)
}
```

### Transition matrix

```{r transition-matrix}
transition_summary <- history_df %>%
  filter(previous_pair != "initial") %>%
  group_by(previous_pair, next_pair) %>%
  summarise(N = n(), .groups = "drop") %>%
  group_by(previous_pair) %>%
  mutate(share = N / sum(N)) %>%
  ungroup() %>%
  mutate(
    previous_pair = label_binary_pair(previous_pair),
    next_pair     = label_binary_pair(next_pair)
  )

knitr::kable(
  transition_summary %>%
    arrange(previous_pair, next_pair) %>%
    mutate(
      N = scales::comma(N),
      share = sprintf("%.1f%%", share * 100)
    ),
  col.names = c("Previous State", "Next State", "N", "Share")
)
```

### Both-Coop Rates by Previous State

```{r both-coop-summary}
history_both_summary <- history_df %>%
  group_by(previous_pair) %>%
  summarise(N = n(), share_both_coop = mean(next_pair == "11"), .groups = "drop") %>%
  bind_rows(
    tibble(previous_pair = "Total", N = nrow(history_df), share_both_coop = mean(history_df$next_pair == "11"))
  ) %>%
  mutate(previous_pair = label_binary_pair(previous_pair)) %>%
  arrange(factor(previous_pair, levels = c("CC", "CD", "DC", "DD", "initial", "Total"))) %>%
  mutate(share_both_coop = sprintf("%.1f%%", share_both_coop * 100), N = scales::comma(N))

knitr::kable(history_both_summary, col.names = c("Previous State", "N", "Share Both Cooperate"))
```

### P1-Coop Rates by Previous State

```{r p1-coop-summary}
history_p1_summary <- history_df %>%
  group_by(previous_pair) %>%
  summarise(N = n(), share_p1_coop = mean(next_p1), .groups = "drop") %>%
  bind_rows(
    tibble(previous_pair = "Total", N = nrow(history_df), share_p1_coop = mean(history_df$next_p1))
  ) %>%
  mutate(previous_pair = label_binary_pair(previous_pair)) %>%
  arrange(factor(previous_pair, levels = c("CC", "CD", "DC", "DD", "initial", "Total"))) %>%
  mutate(share_p1_coop = sprintf("%.1f%%", share_p1_coop * 100), N = scales::comma(N))

knitr::kable(history_p1_summary, col.names = c("Previous State", "N", "Share P1 Cooperate"))
```

## Regressions

### P1 Overall Cooperation Rate \~ Delta_rd + g + l + delta

```{r reg-coop-rate}
p1_coop_rate_model <- lm(
  p1_coop_rate ~ Delta_rd + g + l + delta,
  data = paired_df
)
summary(p1_coop_rate_model)
```

### P1 Round 1 Cooperation \~ Delta_rd + g + l + delta

```{r reg-r1-coop}
p1_r1_coop_model <- lm(
  p1_r1_coop ~ Delta_rd + g + l + delta,
  data = paired_df
)
summary(p1_r1_coop_model)
```

### Next P1 action (linear probability model)

Round-\(t\) P1 cooperation (0/1) from `history_df`: **initial** means the first round of a supergame (no prior joint action); **non-initial** uses joint memory-1 state with **CC** (`11`) as the reference category, plus game parameters and `round_next` (round index within the supergame).

```{r reg-next-p1-initial}
history_initial <- history_df %>% filter(previous_pair == "initial")
next_p1_initial_lpm <- lm(
  next_p1 ~ Delta_rd + g + l + delta,
  data = history_initial
)
summary(next_p1_initial_lpm)
```

```{r reg-next-p1-noninitial}
history_noninitial <- history_df %>%
  filter(previous_pair != "initial") %>%
  mutate(previous_pair = relevel(factor(previous_pair), ref = "11"))
next_p1_noninitial_lpm <- lm(
  next_p1 ~ previous_pair + Delta_rd + g + l + delta + round_next,
  data = history_noninitial
)
summary(next_p1_noninitial_lpm)
```
