---
title: "P1 Next-Action Prediction Results"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Overview

We predict whether P1 cooperates in round $t$ of a supergame using several
model families.  The **baseline** models use game metadata ($g$, $l$,
$\delta$) plus P1's own past-action n-grams (last 1--5 actions, one-hot
encoded).  These are fit with **OLS**, **Lasso**, and **SVR** (RBF
kernel approximation + LinearSVR).

**IRL-SG** is the 6-parameter learning model of Fudenberg \& Karreskog
Rehbinder (2024, Appendix D): a logistic initial-round reinforcement learner
with semi-grim memory-1 transitions at non-initial rounds.  We report two
variants: **IRL-SG (defaults)** uses the paper's published parameters
$(\alpha, \beta, \lambda, p_{CC}, p_{CD/DC}, p_{DD})$, while
**IRL-SG (MLE)** re-estimates those parameters via maximum likelihood on
the training set---as the paper does in its one-step-ahead prediction
exercise.

Train/test split is at the **sequence level** (80/20 of supergame rows) for
the baseline models and at the **session level** (80/20 of sessions) for
IRL-SG. Both are stratified on unique $(g, l, \delta)$ treatments.

```{r load-data}
results_dir <- "../results/next_action_prediction"
csv_files <- list.files(results_dir, pattern = "^next_action_results_model_.*\\.csv$", full.names = TRUE)
results <- bind_rows(lapply(csv_files, read_csv, show_col_types = FALSE)) %>%
  distinct(feature_spec, model, .keep_all = TRUE)

spec_order <- c("meta_only", paste0("meta_hist_", 1:5), "irl_sg", "irl_sg_mle")
spec_labels <- c(
  meta_only   = "g, l, \u03b4 only",
  meta_hist_1 = "+ last 1",
  meta_hist_2 = "+ last 2",
  meta_hist_3 = "+ last 3",
  meta_hist_4 = "+ last 4",
  meta_hist_5 = "+ last 5",
  irl_sg      = "IRL-SG (defaults)",
  irl_sg_mle  = "IRL-SG (MLE)"
)

results <- results %>%
  filter(feature_spec %in% spec_order) %>%
  mutate(
    feature_spec = factor(feature_spec, levels = spec_order),
    feature_label = recode(feature_spec, !!!spec_labels),
    feature_label = factor(feature_label, levels = spec_labels),
    model = case_when(
      model == "irl_sg_mle" ~ "IRL-SG (MLE)",
      model == "irl_sg"     ~ "IRL-SG (defaults)",
      TRUE                  ~ toupper(model)
    )
  )

baseline <- results %>% filter(!grepl("IRL-SG", model))
irl <- results %>% filter(grepl("IRL-SG", model))
```

## Sample sizes

Longer n-gram requirements exclude shorter sequences.

```{r sample-sizes}
results %>%
  distinct(feature_label, model, train_rows, test_rows) %>%
  arrange(feature_label, model) %>%
  mutate(across(c(train_rows, test_rows), scales::comma)) %>%
  select(Features = feature_label, Model = model,
         `Train rows` = train_rows, `Test rows` = test_rows) %>%
  knitr::kable(align = "llrr")
```

## Out-of-sample accuracy

```{r table-accuracy}
results %>%
  select(feature_label, model, test_accuracy) %>%
  pivot_wider(names_from = model, values_from = test_accuracy) %>%
  arrange(feature_label) %>%
  mutate(across(-feature_label, ~ ifelse(is.na(.x), "", sprintf("%.2f%%", .x * 100)))) %>%
  rename(Features = feature_label) %>%
  knitr::kable(align = c("l", rep("r", ncol(.) - 1)))
```

```{r plot-accuracy, fig.width=8, fig.height=4.5}
ggplot(results, aes(x = feature_label, y = test_accuracy, color = model, group = model)) +
  geom_point(size = 3) +
  geom_line(data = baseline) +
  geom_point(data = irl, shape = 18, size = 5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0.6, 1)) +
  labs(x = NULL, y = "Test Accuracy", color = "Model",
       title = "Out-of-sample accuracy by feature set and model") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "top")

ggsave("../figures/prediction_accuracy.png", width = 8, height = 4.5)
```

## Out-of-sample AUC

```{r table-auc}
results %>%
  select(feature_label, model, test_auc) %>%
  pivot_wider(names_from = model, values_from = test_auc) %>%
  arrange(feature_label) %>%
  mutate(across(-feature_label, ~ ifelse(is.na(.x), "", sprintf("%.4f", .x)))) %>%
  rename(Features = feature_label) %>%
  knitr::kable(align = c("l", rep("r", ncol(.) - 1)))
```

```{r plot-auc, fig.width=8, fig.height=4.5}
ggplot(results, aes(x = feature_label, y = test_auc, color = model, group = model)) +
  geom_point(size = 3) +
  geom_line(data = baseline) +
  geom_point(data = irl, shape = 18, size = 5) +
  scale_y_continuous(limits = c(0.5, 1)) +
  labs(x = NULL, y = "Test AUC", color = "Model",
       title = "Out-of-sample AUC by feature set and model") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "top")

ggsave("../figures/prediction_auc.png", width = 8, height = 4.5)
```

## Out-of-sample log loss

```{r table-logloss}
results %>%
  select(feature_label, model, test_log_loss) %>%
  pivot_wider(names_from = model, values_from = test_log_loss) %>%
  arrange(feature_label) %>%
  mutate(across(-feature_label, ~ ifelse(is.na(.x), "", sprintf("%.4f", .x)))) %>%
  rename(Features = feature_label) %>%
  knitr::kable(align = c("l", rep("r", ncol(.) - 1)))
```

```{r plot-logloss, fig.width=8, fig.height=4.5}
ggplot(results, aes(x = feature_label, y = test_log_loss, color = model, group = model)) +
  geom_point(size = 3) +
  geom_line(data = baseline) +
  geom_point(data = irl, shape = 18, size = 5) +
  labs(x = NULL, y = "Test Log Loss", color = "Model",
       title = "Out-of-sample log loss by feature set and model") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "top")

ggsave("../figures/prediction_logloss.png", width = 8, height = 4.5)
```

## Out-of-sample MSE

```{r table-mse}
results %>%
  select(feature_label, model, test_mse) %>%
  pivot_wider(names_from = model, values_from = test_mse) %>%
  arrange(feature_label) %>%
  mutate(across(-feature_label, ~ ifelse(is.na(.x), "", sprintf("%.4f", .x)))) %>%
  rename(Features = feature_label) %>%
  knitr::kable(align = c("l", rep("r", ncol(.) - 1)))
```

```{r plot-mse, fig.width=8, fig.height=4.5}
ggplot(results, aes(x = feature_label, y = test_mse, color = model, group = model)) +
  geom_point(size = 3) +
  geom_line(data = baseline) +
  geom_point(data = irl, shape = 18, size = 5) +
  labs(x = NULL, y = "Test MSE", color = "Model",
       title = "Out-of-sample MSE by feature set and model") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "top")

ggsave("../figures/prediction_mse.png", width = 8, height = 4.5)
```

## In-sample vs out-of-sample (overfitting check)

```{r table-in-vs-out}
results %>%
  select(feature_label, model, train_accuracy, test_accuracy) %>%
  mutate(
    gap = train_accuracy - test_accuracy,
    across(c(train_accuracy, test_accuracy, gap), ~ sprintf("%.2f%%", .x * 100))
  ) %>%
  rename(Features = feature_label, Model = model,
         `Train Acc` = train_accuracy, `Test Acc` = test_accuracy,
         Gap = gap) %>%
  knitr::kable(align = "llrrr")
```

```{r plot-overfit, fig.width=8, fig.height=4.5}
results %>%
  select(feature_label, model, train_accuracy, test_accuracy) %>%
  pivot_longer(c(train_accuracy, test_accuracy), names_to = "split", values_to = "accuracy") %>%
  mutate(split = recode(split, train_accuracy = "Train", test_accuracy = "Test")) %>%
  ggplot(aes(x = feature_label, y = accuracy, color = model, linetype = split, group = interaction(model, split))) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0.6, 1)) +
    labs(x = NULL, y = "Accuracy", color = "Model", linetype = "Split",
         title = "Train vs test accuracy") +
    theme_bw(base_size = 13) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "top")

ggsave("../figures/prediction_overfit.png", width = 8, height = 4.5)
```
