---
title: "Bootstrap Test-Set Prediction Results"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Overview

Every model is fit **once** on the same session-level stratified train set
(seed=0, test_size=0.2). The test set is then resampled with replacement
$B = 1{,}000$ times to form 95% CIs on the held-out metrics.

The results are split into two figures, each keeping its models on an
apples-to-apples evaluation scope:

- **Figure A — all rounds (incl. $t=0$, "no-mask")**: models evaluated on
  every round including the opening action. These have no subject-level
  memory at $t=0$: OLS with only $g, \ell, \delta$ and the Transformer
  trained without round-1 masking.
- **Figure B — rounds $t \ge 1$ only ("mask")**: models evaluated on rounds
  with at least one prior joint action. Includes OLS with joint $n$-gram
  history ($+$ last 1…5), the Transformer trained with round-1 labels
  masked, and IRL-SG (defaults & MLE) scored on the same t≥1 rounds.

IRL-SG does predict $t=0$ internally (using $\sigma(\alpha + \beta \Delta_{RD} +
e_i)$), but its $t=0$ probabilities use an individual-specific experience
term $e_i$ that accumulates across supergames — neither the OLS meta-only
nor the Transformer have access to this. To keep Fig B comparable we score
IRL-SG only on its in-game Markov-1 transitions ($t \ge 1$).

### IRL-SG, briefly

Six-parameter model.
*Initial round* ($t=0$): $P(\text{coop}) = \sigma(\alpha + \beta \Delta_{RD} + e_i)$,
with $e_i$ an individual experience term that updates from the prior
supergame's initial action and total payoff.
*In-game rounds* ($t \ge 1$): three fixed Markov-1 transition probabilities on
the previous joint action — $p_{CC}, p_{CD\_DC}, p_{DD}$ — defining a
"semi-grim" strategy.

Parameters are either taken from the paper (defaults) or fit via
`differential_evolution` MLE on the training set.

## Load data

```{r load-data}
results_dir <- "../results/next_action_prediction"

baseline_files <- list.files(results_dir, pattern = "^bootstrap_baseline_B.*\\.csv$",  full.names = TRUE)
irl_files      <- list.files(results_dir, pattern = "^bootstrap_irl_sg_.*B.*\\.csv$",  full.names = TRUE)
neural_files   <- list.files(results_dir, pattern = "^bootstrap_neural_.*B.*\\.csv$",  full.names = TRUE)

bs_baseline <- bind_rows(lapply(baseline_files, read_csv, show_col_types = FALSE))
bs_irl      <- if (length(irl_files)    > 0) bind_rows(lapply(irl_files,    read_csv, show_col_types = FALSE)) else NULL
bs_neural   <- if (length(neural_files) > 0) bind_rows(lapply(neural_files, read_csv, show_col_types = FALSE)) else NULL

spec_order  <- c("meta_only", paste0("meta_hist_", 1:5))
spec_labels <- c(
  meta_only   = "g, l, \u03b4 only",
  meta_hist_1 = "+ last 1",
  meta_hist_2 = "+ last 2",
  meta_hist_3 = "+ last 3",
  meta_hist_4 = "+ last 4",
  meta_hist_5 = "+ last 5"
)

bs_baseline <- bs_baseline %>%
  filter(feature_spec %in% spec_order) %>%
  mutate(
    feature_spec  = factor(feature_spec, levels = spec_order),
    feature_label = recode(feature_spec, !!!spec_labels),
    feature_label = factor(feature_label, levels = spec_labels),
    model         = toupper(model)
  )

irl_model_map <- c(
  irl_sg             = "IRL-SG (defaults)",
  irl_sg_mle         = "IRL-SG (MLE)",
  irl_sg_mask_r1     = "IRL-SG (defaults)",
  irl_sg_mle_mask_r1 = "IRL-SG (MLE)"
)
neural_model_map <- c(
  transformer_full    = "Transformer (no-mask)",
  transformer_mask_r1 = "Transformer (mask r1)"
)
```

## Sequence length distribution

```{r load-raw}
library(jsonlite)
raw_df <- read_csv("../data/experimental_data.csv", show_col_types = FALSE) %>%
  mutate(seq_len = map_int(c, ~ length(fromJSON(.x))))
```

```{r hist-seqlen, fig.width=8, fig.height=4}
ggplot(raw_df, aes(x = seq_len)) +
  geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7, color = "white") +
  labs(x = "Supergame length (rounds)", y = "Count",
       title = "Distribution of supergame lengths") +
  theme_bw(base_size = 13)

ggsave("../figures/bootstrap_seqlen_hist.png", width = 8, height = 4)
```

```{r seqlen-summary}
tibble(
  Stat = c("Min", "25th pctl", "Median", "Mean", "75th pctl", "95th pctl", "Max"),
  Value = c(
    min(raw_df$seq_len),
    quantile(raw_df$seq_len, 0.25),
    median(raw_df$seq_len),
    sprintf("%.1f", mean(raw_df$seq_len)),
    quantile(raw_df$seq_len, 0.75),
    quantile(raw_df$seq_len, 0.95),
    max(raw_df$seq_len)
  )
) %>% knitr::kable()
```

## Accuracy and AUC by sequence length

Out-of-sample metrics on the held-out session split, comparing the two
Figure A ("no-mask") models that predict every round including $t=0$:
**OLS (g, l, δ)** and **Transformer (no-mask)**. Points show mean
accuracy / AUC per supergame length, weighted by rounds (larger
buckets => larger markers). Red line is a LOESS smoother.

```{r load-perround-preds}
preds_ols <- read_csv(
  "../results/next_action_prediction/test_preds_ols_meta_only_seed_0.csv",
  show_col_types = FALSE
) %>% mutate(model_label = "OLS (g, l, \u03b4)")

preds_trf_path <- "../results/next_action_prediction/test_preds_transformer_full_seed_0.csv"
preds_trf <- if (file.exists(preds_trf_path)) {
  read_csv(preds_trf_path, show_col_types = FALSE) %>%
    mutate(model_label = "Transformer (no-mask)")
} else {
  NULL
}

seq_len_lookup <- raw_df %>%
  transmute(session, id, inter, seq_len)

preds_all <- bind_rows(preds_ols, preds_trf) %>%
  inner_join(seq_len_lookup, by = c("session", "id", "inter"))
```

```{r bucket-by-seqlen}
simple_auc <- function(y, p) {
  if (length(unique(y)) < 2) return(NA_real_)
  pos <- p[y == 1]
  neg <- p[y == 0]
  mean(outer(pos, neg, ">")) + 0.5 * mean(outer(pos, neg, "=="))
}

by_len <- preds_all %>%
  group_by(model_label, seq_len) %>%
  summarise(
    n_rounds = n(),
    accuracy = mean((p >= 0.5) == y),
    auc      = simple_auc(y, p),
    .groups  = "drop"
  ) %>%
  filter(n_rounds >= 50)
```

```{r plot-acc-by-seqlen, fig.width=10, fig.height=4.5}
ggplot(by_len, aes(x = seq_len, y = accuracy, color = model_label)) +
  geom_point(aes(size = n_rounds), alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_size_continuous(labels = scales::comma) +
  scale_color_manual(values = c("OLS (g, l, \u03b4)" = "steelblue",
                                "Transformer (no-mask)" = "firebrick")) +
  labs(x = "Supergame length (rounds)", y = "Accuracy",
       size = "N rounds", color = "Model",
       title = "Out-of-sample accuracy by supergame length") +
  theme_bw(base_size = 13)

ggsave("../figures/bootstrap_acc_by_seqlen.png", width = 10, height = 4.5)
```

```{r plot-auc-by-seqlen, fig.width=10, fig.height=4.5}
ggplot(by_len %>% filter(!is.na(auc)),
       aes(x = seq_len, y = auc, color = model_label)) +
  geom_point(aes(size = n_rounds), alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_size_continuous(labels = scales::comma) +
  scale_color_manual(values = c("OLS (g, l, \u03b4)" = "steelblue",
                                "Transformer (no-mask)" = "firebrick")) +
  labs(x = "Supergame length (rounds)", y = "AUC",
       size = "N rounds", color = "Model",
       title = "Out-of-sample AUC by supergame length") +
  theme_bw(base_size = 13)

ggsave("../figures/bootstrap_auc_by_seqlen.png", width = 10, height = 4.5)
```

## Figure A — all rounds (incl. $t=0$, "no-mask")

Models with no subject-level memory that predict every round, including
the opening action.

```{r build-figA}
figA_models_levels <- c("OLS (g, l, \u03b4)", "Transformer (no-mask)")

figA <- list()

if (nrow(bs_baseline) > 0) {
  figA$ols <- bs_baseline %>%
    filter(feature_spec == "meta_only", model == "OLS") %>%
    mutate(model_label = "OLS (g, l, \u03b4)") %>%
    select(bootstrap_iter, model_label, test_accuracy, test_auc, test_log_loss)
}

if (!is.null(bs_neural)) {
  figA$neu <- bs_neural %>%
    mutate(model_label = recode(model, !!!neural_model_map)) %>%
    filter(model_label == "Transformer (no-mask)") %>%
    select(bootstrap_iter, model_label, test_accuracy, test_auc, test_log_loss)
}

figA_df <- bind_rows(figA) %>%
  mutate(model_label = factor(model_label, levels = figA_models_levels))
```

```{r figA-table}
figA_tbl <- figA_df %>%
  group_by(model_label) %>%
  summarise(
    Accuracy = sprintf("%.2f%% [%.2f, %.2f]",
                       median(test_accuracy) * 100,
                       quantile(test_accuracy, 0.025) * 100,
                       quantile(test_accuracy, 0.975) * 100),
    AUC = sprintf("%.4f [%.4f, %.4f]",
                  median(test_auc),
                  quantile(test_auc, 0.025),
                  quantile(test_auc, 0.975)),
    NLL = sprintf("%.4f [%.4f, %.4f]",
                  median(test_log_loss),
                  quantile(test_log_loss, 0.025),
                  quantile(test_log_loss, 0.975)),
    .groups = "drop"
  ) %>%
  rename(Model = model_label)

knitr::kable(figA_tbl, align = "lrrr", caption = "Figure A: test metrics, all rounds (incl. t=0)")
```

```{r figA-plot, fig.width=9, fig.height=4.5}
figA_ci <- figA_df %>%
  pivot_longer(c(test_accuracy, test_auc, test_log_loss),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric,
    test_accuracy = "Accuracy", test_auc = "AUC", test_log_loss = "NLL")) %>%
  group_by(model_label, metric) %>%
  summarise(lo = quantile(value, 0.025),
            med = median(value),
            hi  = quantile(value, 0.975), .groups = "drop")

ggplot(figA_ci, aes(x = model_label, y = med, color = model_label)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.5) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = NULL, y = "Median + 95% CI",
       title = "Figure A: out-of-sample metrics, all rounds (incl. t=0)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

ggsave("../figures/bootstrap_figA_ci.png", width = 9, height = 4.5)
```

## Figure B — rounds $t \ge 1$ only ("mask")

OLS with joint $n$-gram history ($+$ last 1…5), Transformer with round-1
labels masked, and IRL-SG (defaults & MLE) scored on the same $t \ge 1$
rounds.

```{r build-figB}
figB_ngram_levels <- c(paste0("OLS (", spec_labels[2:6], ")"))
figB_models_levels <- c(figB_ngram_levels, "Transformer (mask r1)",
                        "IRL-SG (defaults)", "IRL-SG (MLE)")

figB <- list()

if (nrow(bs_baseline) > 0) {
  figB$ols <- bs_baseline %>%
    filter(feature_spec %in% paste0("meta_hist_", 1:5), model == "OLS") %>%
    mutate(model_label = paste0("OLS (", as.character(feature_label), ")")) %>%
    select(bootstrap_iter, model_label, test_accuracy, test_auc, test_log_loss)
}

if (!is.null(bs_neural)) {
  figB$neu <- bs_neural %>%
    mutate(model_label = recode(model, !!!neural_model_map)) %>%
    filter(model_label == "Transformer (mask r1)") %>%
    select(bootstrap_iter, model_label, test_accuracy, test_auc, test_log_loss)
}

if (!is.null(bs_irl)) {
  figB$irl <- bs_irl %>%
    filter(model %in% c("irl_sg_mask_r1", "irl_sg_mle_mask_r1")) %>%
    mutate(model_label = recode(model, !!!irl_model_map)) %>%
    select(bootstrap_iter, model_label, test_accuracy, test_auc, test_log_loss)
}

figB_df <- bind_rows(figB) %>%
  mutate(model_label = factor(model_label, levels = figB_models_levels))
```

```{r figB-table}
figB_tbl <- figB_df %>%
  group_by(model_label) %>%
  summarise(
    Accuracy = sprintf("%.2f%% [%.2f, %.2f]",
                       median(test_accuracy) * 100,
                       quantile(test_accuracy, 0.025) * 100,
                       quantile(test_accuracy, 0.975) * 100),
    AUC = sprintf("%.4f [%.4f, %.4f]",
                  median(test_auc),
                  quantile(test_auc, 0.025),
                  quantile(test_auc, 0.975)),
    NLL = sprintf("%.4f [%.4f, %.4f]",
                  median(test_log_loss),
                  quantile(test_log_loss, 0.025),
                  quantile(test_log_loss, 0.975)),
    .groups = "drop"
  ) %>%
  rename(Model = model_label)

knitr::kable(figB_tbl, align = "lrrr", caption = "Figure B: test metrics, rounds t >= 1 only")
```

```{r figB-plot, fig.width=10, fig.height=4.5}
figB_ci <- figB_df %>%
  pivot_longer(c(test_accuracy, test_auc, test_log_loss),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric,
    test_accuracy = "Accuracy", test_auc = "AUC", test_log_loss = "NLL")) %>%
  group_by(model_label, metric) %>%
  summarise(lo = quantile(value, 0.025),
            med = median(value),
            hi  = quantile(value, 0.975), .groups = "drop")

ggplot(figB_ci, aes(x = model_label, y = med, color = model_label)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.5) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = NULL, y = "Median + 95% CI",
       title = "Figure B: out-of-sample metrics, rounds t >= 1 only") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

ggsave("../figures/bootstrap_figB_ci.png", width = 10, height = 4.5)
```

## Pairwise differences: does more history help? (Figure B models)

For each bootstrap iteration we compute the difference in test accuracy
between adjacent OLS $n$-gram specs. If the 95% CI excludes zero, the
improvement is robust.

```{r pairwise-diffs, fig.width=10, fig.height=5}
spec_pairs <- tibble(
  from  = spec_labels[1:5],
  to    = spec_labels[2:6],
  label = paste(spec_labels[2:6], "vs", spec_labels[1:5])
)

diffs <- list()
bs_ols <- bs_baseline %>% filter(model == "OLS")
for (i in seq_len(nrow(spec_pairs))) {
  d <- bs_ols %>%
    filter(feature_label %in% c(spec_pairs$from[i], spec_pairs$to[i])) %>%
    select(bootstrap_iter, feature_label, test_accuracy) %>%
    pivot_wider(names_from = feature_label, values_from = test_accuracy)
  if (ncol(d) < 3) next
  d$diff <- d[[spec_pairs$to[i]]] - d[[spec_pairs$from[i]]]
  diffs[[i]] <- tibble(comparison = spec_pairs$label[i], diff = d$diff)
}
diff_df <- bind_rows(diffs) %>%
  mutate(comparison = factor(comparison, levels = spec_pairs$label))

ggplot(diff_df, aes(x = diff)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ comparison, scales = "free") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(x = "Accuracy difference", y = "Count",
       title = "Pairwise accuracy gains from additional history (OLS)") +
  theme_bw(base_size = 11)

ggsave("../figures/bootstrap_pairwise_diffs.png", width = 10, height = 5)
```

```{r pairwise-ci-table}
diff_ci <- diff_df %>%
  group_by(comparison) %>%
  summarise(
    lo  = quantile(diff, 0.025),
    med = median(diff),
    hi  = quantile(diff, 0.975),
    pct_positive = mean(diff > 0),
    .groups = "drop"
  ) %>%
  mutate(
    CI = sprintf("%.3f%% [%.3f, %.3f]", med * 100, lo * 100, hi * 100),
    `Pr(>0)` = sprintf("%.1f%%", pct_positive * 100)
  ) %>%
  select(Comparison = comparison, `Median diff (CI)` = CI, `Pr(>0)`)

knitr::kable(diff_ci, align = "lrr")
```

## IRL-SG: defaults vs MLE (scoring scopes side by side)

We score the same IRL-SG parameters in two ways: on all rounds (incl.
$t=0$, where the $\sigma(\alpha + \beta \Delta_{RD} + e_i)$ initial-round
logit + subject experience term contribute) and on $t \ge 1$ only (just
the Markov-1 transitions). The gap tells you how much of IRL-SG's
aggregate performance is driven by its initial-round component.

```{r irl-scope-table, eval=!is.null(bs_irl)}
irl_scope <- bs_irl %>%
  mutate(
    scope = if_else(grepl("mask_r1$", model), "t \u2265 1 only", "all rounds (incl. t=0)"),
    variant = case_when(
      grepl("mle", model)  ~ "IRL-SG (MLE)",
      TRUE                 ~ "IRL-SG (defaults)"
    )
  ) %>%
  group_by(variant, scope) %>%
  summarise(
    Accuracy = sprintf("%.2f%% [%.2f, %.2f]",
                       median(test_accuracy) * 100,
                       quantile(test_accuracy, 0.025) * 100,
                       quantile(test_accuracy, 0.975) * 100),
    AUC = sprintf("%.4f [%.4f, %.4f]",
                  median(test_auc),
                  quantile(test_auc, 0.025),
                  quantile(test_auc, 0.975)),
    NLL = sprintf("%.4f [%.4f, %.4f]",
                  median(test_log_loss),
                  quantile(test_log_loss, 0.025),
                  quantile(test_log_loss, 0.975)),
    .groups = "drop"
  )

knitr::kable(irl_scope, align = "llrrr")
```

```{r irl-mle-vs-defaults, eval=!is.null(bs_irl)}
mle_col <- "irl_sg_mle_mask_r1"
def_col <- "irl_sg_mask_r1"
if (mle_col %in% bs_irl$model && def_col %in% bs_irl$model) {
  irl_wide <- bs_irl %>%
    filter(model %in% c(mle_col, def_col)) %>%
    select(bootstrap_iter, model, test_accuracy, test_auc, test_log_loss) %>%
    pivot_wider(names_from = model, values_from = c(test_accuracy, test_auc, test_log_loss))

  qfmt <- function(x) sprintf("%.4f [%.4f, %.4f]", median(x), quantile(x, 0.025), quantile(x, 0.975))

  irl_diff <- tibble(
    Metric = c("Accuracy", "AUC", "NLL"),
    `MLE - Defaults (95% CI), t >= 1` = c(
      qfmt(irl_wide[[paste0("test_accuracy_", mle_col)]] - irl_wide[[paste0("test_accuracy_", def_col)]]),
      qfmt(irl_wide[[paste0("test_auc_", mle_col)]]      - irl_wide[[paste0("test_auc_", def_col)]]),
      qfmt(irl_wide[[paste0("test_log_loss_", mle_col)]] - irl_wide[[paste0("test_log_loss_", def_col)]])
    )
  )
  knitr::kable(irl_diff, align = "lr")
}
```
